Topics in Statistical Theory --- Example Sheet 1

### 2020-10-31 modern_statistical_methods/msm_ES2.tex
progress?

	Using the random feature map $\hat\phi(z) = \psi_{M_z}$ we can approximate kernel ridge regression using the random feature map method. 
\end{proof}
\end{document}

>>>>

	Using the random feature map $\hat\phi(z) = \psi_{M_z}$ we can approximate kernel ridge regression using the random feature map method. 
\end{proof}

\begin{question}
	Consider the logistic regression model where we assume $Y_1, \dotsc, Y_n \in \qty{-1, 1}$ are independent and
	\[
	\log(\frac{\PP(Y_i = 1)}{\PP(Y_i = -1)}) = x_i\T\beta^0. 
	\]
	Show that the maximum likelihood estimate $\beta$ minimises
	\[
	\sum_{i=1}^n \log(1 + \exp(-Y_i x_i\T \beta))
	\]
	over $\beta \in \RR^p$. 
\end{question}

\begin{proof}
	Let $(y_1, \dotsc, y_n)$ be the responses, then the likelihood function becomes
	\[
	L(\beta) = \PP(Y_1, \dotsc, Y_n \mid \beta)
	\]

	Note that
	\begin{align*}
		\frac{\PP(Y_i = 1)}{1 - \PP(Y_i = 1)} = \exp(x_i\T\beta) \implies \PP(Y_i = 1) = \frac{1}{1 + \exp(-x_i\T \beta)},
	\end{align*}
and analogously
\begin{align*}
	\PP(Y_i = -1) =  \frac{1}{1 + \exp(x_i\T\beta)}. 
\end{align*}
We can combine the above formulas as
\[
\PP(Y_i = y_i) = \frac{1}{1 + \exp(-y_i x_i\T\beta)}. 
\]

Therefore our the MLE $\hat\beta$ maximises
\[
L(\beta) = \prod_{i=1}^n \frac{1}{1 + \exp(-Y_i x_i\T\beta)}, 
\]
and it also maximises the log-likelihood function
\[
\log(L(\beta)) = - \sum_{i=1}^n \log(1 + \exp(-Y_i x_i\T\beta))
\]
which is of course equivalent to minimising
\[
-\log(L(\beta)) = \sum_{i=1}^n \log(1 + \exp(-Y_i x_i\T\beta)). 
\]
\end{proof}

\begin{question}
	Consider the following algorithm for model selection when we have a response $Y \in \RR^n$ and a matrix of predictors $X \in \RR^{n \times p}$.
	\begin{enumerate}[(a)]
		\item First centre $Y$ and all the columns of $X$. Initiale the current model $M \subseteq \qty{1, \dotsc, p}$ to be $\emptyset$ and set the current residual $R$ to be $Y$. 
		\item Find the variable $k^*$ in $M^c$ most correlated with the current residual $R$. Set $M$ to be $M \cup \qty{k^*}$. Replace $R$ with the residual from regressing $R$ onto $X_{k^*}$. Further replac eeach variable in $M^c$ with the residual from regressing itself onto $X_{k^*}$.
		\item Continue the previous step unto $R = 0$. 
	\end{enumerate}
Show that this algorithm is equivalent to forward selection.

\emph{Hint:} Use induction on the iteration $m$ of the algorithm. Consider strengthening the natural induction hypothesis that the model at iteration $m$ is the same as that selected after $m$ steps of forward selection. 
\end{question}

\begin{solution}
	content...
\end{solution}

\begin{question}
	Show that if $W$ is mean-zero and sub-Gaussian with parameter $\sigma$, then $\Var(W) \leq \sigma^2$. 
\end{question}

\begin{proof}
	If $W$ is mean-zero, then $\Var(W) = \EE(W^2)$. By the proof of lemma 15 we have $\EE(W^2) \leq 2 \sigma^2$, but this bound is too loose.

	Since $W$ is sub-Gaussian, we have for all $\alpha \in \RR$ that
	\begin{align*}
		\EE[e{\alpha W}] &\leq e^{\alpha^2\sigma^2/2} \\
		\sum_{k=0}^\infty \frac{\EE[W^k] }{k!} \alpha^k &\leq \sum_{k=0}^\infty \frac{\sigma^{2k}}{2^k k!} \alpha^{2k} \\
		\frac{\EE[W^2]}{2} \alpha^2 + \sum_{k=3}^\infty \frac{\EE[W^k]}{k!} \alpha^k &\leq \frac{\sigma^2}{2} \alpha^2 + \sum_{k=2}^\infty \frac{\sigma^{2k}}{2^k k!} \alpha^{2k} \\
		\frac12\alpha^2 \qty(\sigma^2 - \EE[W^2]) &\geq \alpha^3 P(\alpha),
		\end{align*}
	where $P$ is a power series in $\alpha$. Rescaling we find \[
	\frac12(\sigma^2 - \EE[W^2]) \geq \alpha P(\alpha),
	\]
	and letting $\alpha \to 0$ also lets $\alpha P(\alpha) \to 0$, and therefore $\sigma^2 - \EE[W^2] \geq 0$ so $\Var(W) \leq \sigma^2$. 
\end{proof}

\begin{question}
	Verify Hoeffding's lemma for the special case where $W$ is a Rademacher random variable, so $W$ takes the values $-1, 1$ each with probability $1/2$. 
\end{question}

\begin{proof}
		If $W$ is a Rademacher random variable then we have, using $(2k)! \geq 2^k k!$ for each $k \in \NN$, that
	\begin{align*}
		\EE[e^{\alpha W}] = \frac12 e^{-\alpha} + \frac12 e^{\alpha} = \sum_{k=0}^\infty \frac{\alpha^{2k}}{(2k)!} \leq \sum_{k=0}^\infty \frac{\alpha^{2k}}{2^k k!} = e^{\alpha^2/2}
	\end{align*}
which shows that $W$ is sub-Gaussian with parameter $1 = (1 -(-1))/2$, so Hoeffding's lemma holds indeed. 
\end{proof}

\begin{question}
	\begin{enumerate}[(a)]
		\item Let $W \sim \chi_d^2$. Show that
		\[
		\PP(\abs{W/d - 1} \geq t) \leq 2e^{-dt^2/8}
		\]
		for $t \in (0, 1)$. You may use the facts that the mgf of a $\chi_1^2$ random variable is $(1 - 2\alpha)^{-1/2}$ for $\alpha < 1/2$, and $e^{-\alpha} (1 - 2\alpha)^{-1/2} \leq e^{2\alpha^2}$ when $\abs{\alpha} < 1/4$.

		\item Let $A \in \RR^{d\times p}$ have i.i.d.\ standard normal entries. Fix $u \in \RR^p$. Use the result above to conclude that
		\[
		\PP\qty(\abs{\frac{\norm{Au}_2^2}{d \norm{u}_2^2} - 1} \geq t) \leq 2e^{-dt^2/8}. 
		\]

		\item Suppose we have data $u_1, \dotsc, u_n \in \RR^p$, with $p$ large and $n \geq 2$. Show that for a given $\eps \in (0, 1)$ and $d > 16\log(n/\sqrt\eps)/t^2$, each data point may be compressed down to $u_i \mapsto Au_i/\sqrt d = w_i$, whilst approximately preserving the distance between the points:
		\[
		\PP\qty(1 - t \leq \frac{\norm{w_i - w_j}_2^2}{\norm{u_i - u_j}_2^2} \leq 1 + t \text{ for all } i \neq j \in \qty{1, \dotsc, n}) \geq 1 - \eps. 
		\]
		This is the famous Johnson-Lindenstrauss Lemma. 
	\end{enumerate}
\end{question}

\begin{proof}
	\begin{enumerate}[(a)]
		\item 
	\end{enumerate}
\end{proof}
\end{document}

### 2020-10-31 topics_statistical_theory/topics_ES1.tex
progress?

% Functions that appear frequently
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Span}{Span}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator\Exp{Exp}
\DeclareMathOperator\Beta{Beta}
\DeclareMathOperator\Gam{Gamma}
\DeclareMathOperator\Cov{Cov}

@@

	Then, defining $t_0 = -\infty$ and $t_{k+1} = \infty$,  we find that
	\begin{gather*}
	\PP\qty[ n \qty(\hat F_n(t_1), \dotsc, 1 - \hat F_n(t_k)) (a_1, \dotsc, a_{n+1})] \\ = \PP[\text{exactly $a_i$ of the $X_i$ lie in $(t_{i-1}, t_i]$ for $i = 1, \dotsc, n$}].
	\end{gather*}

This probability is 0 unless all $a_i$ are nonnegative integers with $\sum_i a_i = n$. In this case, we can first choose $a_1$ of the $X_i$ which must lie in $(\infty, t_1]$, then we choose $a_2$ of the remaining $X_i$ which must lie in $(t_1, t_2]$ and so forth. We find that the above probability equals
\[
\binom{n}{a_1} (F(t_1))^{a_1} \binom{n - a_1}{a_2} (F(t_2) - F(t_1))^{a_2} \dotsb \binom{a_k + a_{k_1}}{a_k} (F(t_k) - F(t_{k-1}))^{a_k} \cdot (1 - F(t_k))^{a_{k+1}}. 
\]

@@

Now we must compute the covariances: assume $s < t$, then  
\begin{align*}
	\Cov(\hat F_n(s), \hat F_n(t)) &= \EE[\hat F_n(s) \hat F_n(t)] - \EE[\hat F_n(s)] \EE[\hat F_n(t)] \\
	&= \frac1{n^2} \sum_{i, j} \EE[\ind_{X_i \leq s} \ind_{X_j \leq t}] - F(s)F(t) \\
	&= \frac1{n^2} \qty(n F(s) + n(n-1) F(s)F(t)) - F(s)F(t) \\
	&= \frac{F(s) + (n-1) F(s) F(t)}{n} - F(s)F(t) = \frac{F(s) - F(s)F(t)}{n}. 
\end{align*}
This gives the diagonal entries $\Sigma_{ij} = \frac{F(t_i) - F(t_i)F(t_j)}{n}$ for $i < j$. 
In the end, we find
\[
\Sigma_{ij} = \frac1n \cdot \begin{cases}
	F(t_i)(1 - F(t_i)) &\text{if $i = j$}, \\
	F(t_{\min(i, j)}) - F(t_i) F(t_j) &\text{if $i \neq j$}. 
\end{cases}
\]
\end{enumerate} 
\end{solution}
\end{document}

>>>>

% Functions that appear frequently
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Span}{Span}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator\Exp{Exp}
\DeclareMathOperator\Beta{Beta}
\DeclareMathOperator\Gam{Gamma}
\DeclareMathOperator\Cov{Cov}
\DeclareMathOperator\diag{diag}

@@

	Then, defining $t_0 = -\infty$ and $t_{k+1} = \infty$,  we find that
	\begin{gather*}
	\PP\qty[ n \qty(\hat F_n(t_1), \dotsc, 1 - \hat F_n(t_k)) = (a_1, \dotsc, a_{k+1})] \\ = \PP[\text{exactly $a_i$ of the $X_i$ lie in $(t_{i-1}, t_i]$ for $i = 1, \dotsc, n$}].
	\end{gather*}

In this case, we have a multinomial distribution with $n$ trials and probabilities $F(t_1), F(t_2) - F(t_1), \dotsc, F(t_k) - F(t_{k-1}), 1 - F(t_k)$. Therefore, the probability is 0 if $\sum_i a_i \neq n$ and else it is
\[
\frac{n!}{a_1! \dotsb a_{k+1}!} F(t_1)^{a_1} \dotsb (1 - F(t_k))^{a_{k+1}}. 
\]

@@

Now we must compute the covariances: assume $s < t$, then  
\begin{align*}
	\Cov(\hat F_n(s), \hat F_n(t)) &= \EE[\hat F_n(s) \hat F_n(t)] - \EE[\hat F_n(s)] \EE[\hat F_n(t)] \\
	&= \frac1{n^2} \sum_{i, j} \EE[\ind_{X_i \leq s} \ind_{X_j \leq t}] - F(s)F(t) \\
	&= \frac1{n^2} \qty(n F(s) + n(n-1) F(s)F(t)) - F(s)F(t) \\
	&= \frac{F(s) + (n-1) F(s) F(t)}{n} - F(s)F(t) = \frac{F(s) - F(s)F(t)}{n}. 
\end{align*}
This gives the diagonal entries $\Sigma_{ij} = \frac{F(t_i) - F(t_i)F(t_j)}{n}$ for $i < j$. 
In the end, we find
\[
\Sigma_{ij} = \frac1n \cdot \begin{cases}
	F(t_i)(1 - F(t_i)) &\text{if $i = j$}, \\
	F(t_{\min(i, j)}) - F(t_i) F(t_j) &\text{if $i \neq j$}. 
\end{cases}
\]
\end{enumerate} 
\end{solution}

\begin{question}
	We say that a continuous process $(B_t)_{t \in [0, 1]}$ is a \emph{standard Brownian motion} on $[0, 1]$ if $B_0 = 0$ and if, for $0 \leq s_1 \leq t_1 \leq \dotsb \leq s_k \leq t_k \leq 1$, we have $(B_{t_1} - B_{s_1}, \dotsb, B_{t_k} - B_{s_k}) \sim N_k(0, \Sigma)$, where $\Sigma \ceq \diag(t_ 1- s_1, \dotsb, t_k - s_k)$. The process $(W_t)_{t \in [0, 1]}$ defined by $W_t \ceq B_t - tB_1$ is called a \emph{Brownian bridge}, or \emph{tied-down Brownian motion}, because $W_0 = W_1 = 0$. Compute the distribution of $(W_{t_1}, \dotsc, W_{t_k})$. 
\end{question}

\begin{solution}
	Note that $W_t = B_t - t B_1 = (1-t) (B_t - B_0) - t (B_1 - B_t)$. Now, since $(B_t - B_0)$ and $(B_1 - B_t)$ are independent with distributions $N(0, t)$ and $N(0, 1-t)$ distributions respectively, we find that 
	\[
	W_t \sim (1-t) N(0, t) + t N(0, 1 - t) = N\qty(0, \frac{t}{\sqrt{1 - t}}) + N\qty(0, \frac{1-t}{\sqrt t}) = N\qty(0, \frac{t}{\sqrt{1 - t}} + \frac{1-t}{\sqrt t}). 
	\]

	??
\end{solution}

\begin{question}
	Let $\phi$ denote the standard normal density function, which is a bounded, second-order kernel. For $r \in \NN_0$, define the $r$-th Hermite polynomial $H_r$ by $H_r(x) \ceq (-1)^r \phi^{(r)}(x) / \phi(x)$. Prove that $H_r$ is a monic polynomial of degree $r$ that is even if $r$ is even and odd if $r$ is odd. Show further that
	\[
	\int_{-\infty}^\infty H_r(u) H_s(u) \phi(u) \dd{u} = \begin{cases}
		(2\pi)^{1/2} r!, & r=s, \\ 0 , & r \neq s.
	\end{cases}
	\]
	Now fix an integer $\ell \geq 2$ and define 
	\[
	K_\ell(u) \ceq \sum_{r=0}^{\ell -1 } \frac{H_r(0) H_r(u)}{(2\pi)^{1/2} r!} e^{-u^2/2}. 
	\]
	Prove that $K_\ell$ is a bounded kernel of order $\ell$. 
\end{question}

\begin{proof}
 	We prove this by induction on $r$. For $r = 0$, we have $H_0(x) = 1$, which is indeed an even monic polynomial of degree 0. 
 	Now, suppose the claim holds for a given $r$, that is, $H_r(x) = (-1)^r \phi^{(r)}(x)/\phi(x) = p(x)$ for some monic polynomial $p$ of degree $r$, which is even if $r$ is even and odd if $r$ is odd. Then we have
 	\begin{align*}
 	\phi^{(r)}(x) &= (-1)^r p(x) \phi(x) = (-1)^r (2\pi)^{-1/2} p(x) \exp(-x^2/2) \\
 	\phi^{(r+1)}(x) &= (-1)^r 2\pi^{-1/2} \qty(p'(x) - x p(x)) \exp(-x^2/2) \\
 	H_{r+1 }(x) &= (-1)^r (-1)^{r+1} \qty(p'(x) - x p(x)) = xp(x) - p'(x). 
 	\end{align*}
 	Now it is clear that $H$ is a monic polynomial of degree $r$ since $p$ was assumed monic. Furthermore, since derivatives of even functions are odd and vice versa, it is clear that $H$ is odd if $p$ is even and vice versa.

	Now, suppose $r < s$, then 
	\begin{align*}
		\int_{-\infty}^\infty H_r(u) H_s(u) \phi(u) \dd{u} &= (-1)^s \int_{-\infty}^\infty H_r(u) \phi^{(s)}(x) \dd{u} \overset{\text{IBP}}= \int_{-\infty}^\infty H_r^{(s)}(u) \phi(u) \dd{u} = 0,
	\end{align*}
since $H_r^{(s)} = 0$ if $r < s$.

However, if $r = s$, then following the same line of reasoning as above and using the fact that $H_r^{(r)} = r!$, we find
\[
\int_{-\infty}^\infty H_r^2(u) \phi(u) \dd{u} = r! \int_{-\infty}^\infty \phi(u) \dd{u} = r!. 
\]

Now we consider $K_\ell$: we have
\[
\int_{-\infty}^\infty K_\ell(u) \dd{u} = \sum_{r=0}^{\ell - 1} \frac{H_r(0)}{r!} \int_{-\infty}^\infty H_r(u) \phi(u) \dd{u} =  \sum_{r=0}^{\ell - 1} \frac{(-1)^r H_r(0)}{r!} \int_{-\infty}^\infty \phi^{(r)}(u) \dd{u}.
\]
Note that every term in the above sum vanishes except for the $r = 0$ term due to the integral, and the $r = 0$ term is 1, so $K_\ell$ is indeed a kernel.

We verify that $K_\ell$ has order $\ell$: let $j \in \qty{1, \dotsc, \ell - 1}$, then we have
\begin{align*}
	\int_{-\infty}^\infty u^j K_\ell(u) \dd{u} = \sum_{r=0}^{\ell - 1} \frac{(-1)^r H_r(0)}{r!} \int_{-\infty}^\infty u^j \phi^{(r)}(u) \dd{u}.
\end{align*}
Clearly, if $j < r$, integral in the sum vanishes. If $j = r$, the integral yields $r!$. So we have
\[
\int_{-\infty}^\infty u^j K_\ell(u) \dd{u} = (-1)^j H_j(0) + \sum_{r=0}^{j-1} \int_{-\infty}^\infty u^j \phi^{(r)}(u) \dd{u}. 
\]
\end{proof}
\end{document}
